### 2021-03-04 paper.Rmd
finally working on paper again

---
title             : "A bicriterion extension of k-means anticlustering"
shorttitle        : "Bicriterion K-means anticlustering"

@@

abstract: |
    Anticlustering is an automated method to partition data sets into groups 
    in such a way that high between-group similarity and high within-group
    heterogeneity is obtained. Anticlustering has important applications in 
    psychological research, for example to partition stimulus sets into 
    equivalent sets, or to divide a group of students into heterogeneous 
    classes. K-means anticlustering is maximizes the error sum of squares, 
    i.e., the sum of squared Euclidean distances from each element to the 
    center of the group to which it belongs. Thereby, k-means anticlustering 
    directly maximizes the similarity of the mean values of numeric 
    attributes across groups, but ignores the spread of the data. 
    This paper presents a straight-forward extension of 
    the k-means objective function that allows to simultaneously maximize 
    similarity with regard to the mean as well as the variance across 
    groups. It is shown that when both criteria receive equal weights, the 
    bicriterion optimization problem reduces to an augmentation of the input 
    data, leaving the original k-means criterion unchanged. Tests on 
    simulated and real data show that similarity with regard to means and 
    variance criteria can usually be achieved to an astonishing degree with 
    almost no cost to either criterion. It is therefore concluded that the 
    extended model should be preferred to classical k-means anticlustering 
    for optimizing between-group similarity. An exception is given for 
    binary data that does not profit from considering equality in 
    variances.

@@

keywords          : "Anticlustering, variance, k-means, k-variance, bicriterion optimization"
wordcount         : "X"

@@

A different approach to anticlustering directly works with feature based 
data and does not require a transformation into pairwise 
dissimilarities. Instead, the sum of the squared Euclidean distances 
between elements and the cluster they are assigned to is minimized 
[@brusco2006branch], most prominently using the k-means heuristic 
[@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed 
k-means objective that is, maximizing the sum of distances between 
elements and cluster centroids -- a data set is partitioned into similar 
groups, thereby reversing the logic of classical clustering. They 
independently from each other coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean 
distances from each element to the center of the cluster to which it 
belongs. As @spath1986 noted, k-means anticlustering thereby directly 
minimizes differences with regard to the means of the numeric variables 
between groups. Hence, groups only become similar with regard to the 
mean of the distribution of the numeric attributes, which may perform 
poorly as a measure of overall similarity; the spread of the data is not 
targeted and may differ between sets. @papenberg2020 showed that 
maximizing the diversity (based on the pairwise Euclidean distances) is 
better suited to minimize difference with regard to both the mean and 
the variance of the data; it was shown that even an entirely random 
split usually resulted in more similar standard deviations as k-means 
solely focuses on the mean. When requiring that groups are overall 
similar to each other, this focus may be detrimental as similar means 
can be obtained even when the underlying distributions are very 
different [@anscombe1973].

In this paper, I present a simple extension of the k-means objective 
function to maximize simililarity with regard to means and variances at 
the same time. The aim is to establish groups that are overall similar 
to each other. It is shown that this is achieved by formulating a 
bicriterion optimization objective for k-means anticlustering. 
Interestingly, under specific---and important---circumstances, this 
reformulation reduces to an augmentation of the original data matrix 
while the optimization objective remains unchanged. This insight has 
important consequences as it allows to adapt standard optimization 
schemes for single criterion anticlustering. I show that the extended 
k-means criterion outperforms other anticlustering criteria when the 
goal is the minimize difference with regard to the mean and variance of 
the different groups. In most cases, both objectives can be optimized
satisfactorily at the same time without any loss on either; when the 
data set is large enough (usually, when $N > XX$), the bicriterion 
optimization leads to > 99.99% objective performance as compared to the 
single optimization of either k-means or k-variance.

@@

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his comprehensive synthesis of half a century of research on 
k-means. K-means clustering is used to partition two-way, two-mode 
data. That is, $N$ elements each having $P$ attributes into $K$ groups 
$(C_1., \ldots, C_K)$, $C_k$ being the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$
$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

In anticlustering applications, we usually impose restrictions on 
the group sizes $n_k$, with the most common restriction being that all 
groups have equal size:

@@

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), also known as the 
within-group variance [@spath1986]. The SSE is the sum of the squared Euclidean 
distances between each data point and its cluster centroid; the *k*'th centroid 
is a vector of length $P$, where each entry is the mean value of one of the $P$ 
attributes across the observations belonging the *k*'th cluster ($k = 1, \ldots, 
K$). The centroid of the *k*'th cluster has the notation 
$\overline{\mathbf{x}}^{(k)}$ with
$$
\overline{\mathbf{x}}^{(k)} = 
  (\overline{x}_1^{(k)}, \overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})
$$

The mean $\overline{x}_j ^{(k)}$ of the *j*th variable on the *k*th cluster is 
computed as:

@@

Based on the original data and the cluster centroids, the SSE---i.e., the k-means 
objective function---is given as:

@@

If the $SSE$ is maximal, the cluster centroids are as close as possible to the 
overall centroid $\overline{\mathbf{x}}$ and therefore to each other 
[@spath1986]. Thus, k-means anticlustering directly optimizes the 
similarity of the mean values of the $P$ attributes across the $K$ clusters.

## Bi-criterion extension

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. In the following, a rather 
straight-forward extension of the SSE criterion is formulated that can be used 
to minimize differences with regard to the variances of the input variables at 
the same time.

For the sake of simplicity, we first consider a set of $N$ one-dimensional 
observations $(x_1, \ldots, x_N)$. The extension to the general case of $P$ 
attributes will be straight forward, because the formulation of the new 
criterion effectively reduces to the generation of new variables, leaving the 
k-means criterion SSE per sé unchanged.

The sample variance of this set of 
observations is computed as the mean of the squared distances between 
observations and their mean $\mathbf{x}$:

@@

Note at this point that the sample variance is basically defined as a mean 
value. We can use this insight and employ k-means anticlustering to equalize 
this *mean* across groups. For a given $K$, groups can be created that are 
similar with regard to the means of $y$---i.e., the variance of $x$---by 
maximizing the SSE, in this one-dimensional case given as:

$$
\mathit{SSE} = 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{i} - \overline{y} ^{(k)})^2
$$

## Weighted-sum approach

$$
\mathit{SSE_{comb}} = 
  w_1
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  w_2
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2
$$

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of 
equalizing means and variances, respectively.

## Unweighted case

Some simplifications can be arranged when ignoring the weights on the two 
criteria.

$$
\mathit{SSE_{comb}} = 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (x_{i} - \overline{x} ^{(k)})^2 + 
  \sum\limits_{k=1}^{K} 
    \sum\limits_{i \in C_k}
    (y_{i} - \overline{y} ^{(k)})^2 \\
  = \sum\limits_{k=1}^{K} 
     \sum\limits_{i \in C_k}
     ((x_{i} - \overline{x} ^{(k)})^2 + 
     (y_{i} - \overline{y} ^{(k)})^2)
$$

When defining $z = (x, y)$, the SSE reduces to the standard case in (X):

$$
\mathit{SSE} = 
  \sum\limits_{j=1}^{2} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$

This logic can be extended to the general case of $P$ attributes; for each 
attribute, a new variable is created as described above, and the entirety of 
all $2P$ variables is then used as input to standard k-means anticlustering.

@@

A motivating example illustrates that aiming to simultaneously optimize 
similarity of means and variances at the same time is not only 
desirable, but may also be feasible. I created 14 data points from a 
univariate normal distribution ($M = 15$, $SD = 3$) and generated all `r 
n_partitions(14, 2)` possible ways to partition the 14 data points into 
$K = 2$ equal-sized groups. I computed the difference in the two means 
and two variances for each partition. This simple difference metric was 
chosen for purposes of interpretability; in fact, there is a a 1-to-1 
correspondence between these differences and $SSE$ and 
$SSE_{\mathrm{VAR}}$, respectively (Figure 1).[^onlyonegroup] Figure 2 
illustrates similarity in means plotted against similarity in variances 
across all partitions. Ideally, a partition is chosen that minimizes 
the difference with regard to means and variances (see the triangular 
symbol). However, optimal k-means anticlustering would create two groups 
that are not particularly similar with regard to their variance and 
chose the partition that is represented by the squared symbol. The 
diamond symbol represents that partition that where the two groups have 
the least difference in variances. The black symbols illustrate the 
other members of the optimal pareto set.

@@

```{r, fig.cap = "Illustrates the 1-to-1 relationship between SSE and difference in means (Panel A) and variance (Panel B). Each data point illustrates a partition, i.e., a way to divide a data set ($N = 14$) into two groups. In Panel C, the relationship between the two criteria across partitions is plotted. The highlighted symbols indicate membership of the pareto set. The partition represented by the diamond minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the triangular minimizes the sum of the differences in means and variances. ", fig.height = 7, fig.width = 6}

# Plot relationship between objectives
layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE), heights = c(1, 1.4))
plot(
  df[, c("diff_mean", "kmeans")],
  xlab = "mean(f1) - mean(f2)",
  ylab = "SSE",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("bottomleft", legend = "A", cex = 1.3, bty = "n")
plot(
  df[, c("diff_var", "kvar")],
  xlab = "var(f1) - var(f2)",
  ylab = "SSE_var",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("bottomleft", legend = "B", cex = 1.3, bty = "n")

par(mar = c(7, 10, 0.4, 8) + 0.1)

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "kmeans",
  ylab = "kvar"
)
legend("bottomleft", legend = "C", cex = 1.3, bty = "n")

>>>>

---
title             : "k-plus: a bicriterion extension for k-means anticlustering"
shorttitle        : "k-plus anticlustering"

@@

abstract: |

    Anticlustering is a method to partition data sets into groups 
    in such a way that high between-group similarity and high within-group 
    heterogeneity is obtained. In psychological research, anticlustering is for example 
    used to to divide stimulus sets into equivalent sets, to split tests into parallel 
    halfs, or to divide a group of students into heterogeneous classes. Different 
    objective functions can be optimized in anticlustering applications. K-means 
    anticlustering maximizes the error sum of squares, i.e., the sum of squared 
    Euclidean distances from each element to the center of the group to which it 
    belongs. K-means anticlustering thereby directly maximizes the similarity of 
    the mean values of numeric attributes across groups, but it ignores the spread of 
    the data. This paper presents k-plus anticlustering, a straight-forward 
    extension of the k-means objective function that allows to simultaneously 
    maximize similarity with regard to the mean as well as the variance across 
    groups. It is shown that the bicriterion optimization problem reduces to an 
    augmentation of the input data, leaving the original k-means criterion 
    unchanged. Tests on simulated and real data show that similarity with regard to 
    means and variance criteria can usually be achieved to an astonishing degree 
    with almost no cost to either criterion. It is therefore concluded that the 
    extended model should be preferred to classical k-means anticlustering for 
    optimizing between-group similarity. An exception is given for binary data that 
    does not profit from considering equality in variances.

@@

keywords          : "Anticlustering, k-means, k-plus, variance, bicriterion optimization"
wordcount         : "X"

@@

A different approach to anticlustering directly works with feature based data 
and does not require a transformation into pairwise dissimilarities. Instead, 
the sum of the squared Euclidean distances between elements and the cluster they 
are assigned to is minimized [@brusco2006branch], most prominently using the 
k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed k-means 
objective that is, maximizing the sum of distances between elements and cluster 
centroids -- a data set is partitioned into similar groups, thereby reversing 
the logic of classical clustering. They coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean distances 
from each element to the center of the cluster to which it belongs. As 
@spath1986 noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, groups 
only become similar with regard to the mean of the distribution of the numeric 
attributes, which may perform poorly as a measure of overall similarity; the 
spread of the data is not targeted and may differ between sets. @papenberg2020 
showed that maximizing the diversity (based on the pairwise Euclidean distances) 
is better suited to minimize difference with regard to both the mean and the 
variance of the data; it was shown that even an entirely random split usually 
resulted in more similar standard deviations as k-means solely focuses on the 
mean. When requiring that groups are overall similar to each other, this focus 
may be misguided as similar means can be obtained even when the underlying 
distributions are different [@anscombe1973].

In this paper, I present k-plus anticlustering, a simple extension of the 
k-means objective function to maximize similarity with regard to means and 
variances at the same time. The aim is to establish groups that are overall 
similar to each other. It is shown that this is achieved by formulating a 
bicriterion optimization objective for k-means anticlustering. Interestingly, 
this reformulation reduces to an augmentation of the original data matrix while 
the optimization objective remains unchanged. This insight has important 
consequences as it allows to adapt standard optimization schemes for single 
criterion anticlustering. I show that the k-plus criterion outperforms other 
anticlustering criteria when the goal is the minimize difference with regard to 
the mean and variance of the different groups. In most cases, both objectives 
can be optimized satisfactorily at the same time without any loss on either; 
when the data set is large enough (usually, when $N > XX$), the bicriterion 
optimization leads to > 99.99% objective performance as compared to the single 
optimization of either k-means or k-variance.

@@

For the purpose of problem formalization, I adopt the notation Steinley (2006) 
provided in his comprehensive synthesis of half a century of research on 
k-means. K-means clustering is used to partition two-way, two-mode data into 
groups. That is, $N$ elements each having $P$ attributes are assigned to $K$ 
groups $(C_1., \ldots, C_K)$, $C_k$ being the set of $n_k$ objects in group $k$. 
The partitioning is subject to the requirements that each element is assigned to 
exactly one group, formalized by the two constraints

$$
\bigcup\limits_{k = 1}^{K} C_j = X
$$

$$
C_j \cap C_k = \emptyset, \; \forall j, k \in \{1, \ldots, K\}, \; j \ne k
$$

In anticlustering applications, we  impose an additional constraint on the group 
sizes $n_k$, where the most common one requires that all groups have equal size:

@@

For a data matrix $\mathbf{X}_{N \times P} = \{x_{ij}\}_{N \times P}$, k-means 
clustering aims to minimize the error sum of squares (SSE), which is the sum of 
the squared Euclidean distances between each data point and its cluster 
centroid:

The *k*'th centroid $\overline{\mathbf{x}}^{(k)} = (\overline{x}_1^{(k)}, 
\overline{x}_2^{(k)} \ldots, \overline{x}_P^{(k)})$ is a vector of length $P$, 
where each entry is the mean value of one of the $P$ attributes, computed across 
all observations belonging the $k$th cluster $C_k$ ($k = 1, \ldots, K$):

@@

@@

If the $SSE$ is maximal, the cluster centroids are as close as possible to the 
overall centroid $\overline{\mathbf{x}}$ and therefore to each other 
[@spath1986]. Thus, k-means anticlustering directly optimizes the 
similarity of the mean attribute values across clusters.

## A criterion to quantify differences in variances

K-means anticlustering---i.e., maximizing the SSE---minimizes differences 
between the means of the input variables. I now present a rather 
straight-forward adaption of the SSE to capture differences in variance between 
groups. To motivate the new criterion, we first consider a set of 
one-dimensional observations $(x_1, \ldots, x_N)$. The extension to the general 
case of $P$ attributes will be straight forward

The sample variance of a set of observations $x = (x_1, \ldots, x_N)$ is 
computed as the mean of the squared distances between observations and their 
mean $\overline{\mathbf{x}}$:

@@

Since the sample variance is basically defined as an average, k-means 
anticlustering can be employed to equalize the variance across groups. That is, 
we can formulate an adaption of the SSE that quantifies the degree to which 
variances differ between groups:

$$
\mathit{SSE_{Var}} = 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{i} - \overline{y} ^{(k)})^2
$$

Maximizing $\mathit{SSE_{Var}}$ will equalize the variance between groups on one 
attribute. The criterion is extended to the general case of $P$ attributes as

$$
\mathit{SSE_{Var}} = 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{y}_j ^{(k)})^2
$$

where

$$
\overline{y}_j ^{(k)} = \frac{1}{n_k} \sum\limits_{i \in C_k} y_{ij}
$$

and

$$
y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2.
$$

Note that $\mathit{SSE_{Var}}$ has the same form as the standard $\mathit{SSE}$. 
However, the input data reflect the squared difference between each data point 
and the average values instead of the raw data points.

## A bicriterion model: k-plus anticlustering

In most applications, it will not be of interest to only maximize $SSE_{Var}$ 
(i.e., equalizing variances) without also considering the standard $SSE$ (i.e., 
equalizing means). A combination of $SSE_{Var}$ and $SSE$, however, can be 
employed to simultaneously optimize both objectives. The most simple approach to 
such a bicriterion optimization is the weighted sum approach [@naidu2014]. That 
is, both criteria are computed independently from each other and then combined 
into a single criterion through a weighted sum:

$$
\mathit{SSE_{kplus}} = w_1 
  \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (y_{ij} - \overline{x}_j ^{(k)})^2 + 
  w_2 \sum\limits_{j=1}^{P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (x_{ij} - \overline{y}_j ^{(k)})^2
$$

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of 
equalizing means and variances, respectively. Often, they are chosen such that 
$w_1 + w_2 = 1$ [@brusco2019].

Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$ 
(i.e., or setting them to the same value). When defining $z_i = (x_{i1}, \ldots, 
x_{iP}, y_{i1}, \ldots, x_{iP})$, i.e., the concatenation of the $x_{ij}$ and 
$y_{ij}$ values, the bicriterion $\mathit{SSE_{kplus}}$ is computed as

$$
\mathit{SSE_{kplus}} =
  \sum\limits_{j=1}^{2 P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2
$$

This formula again has the same form as SSE (and $SSE_{Var}$), but is computed 
on different data. In this case, both the original data $x_{ij}$ as well as the 
squared distances $y_{ij}$ are used as attributes that enter the objective 
function. That is, computing the k-plus criterion is done by first computing the 
values $y_{ij}$, appending these to the original data matrix, and then compute 
SSE on the augmented data matrix.

Note that a minor problem arises when ignoring the weights during optimizing the 
k-plus objective. Since their computation involves squaring, the squared 
distances $y_{ij}$ have much larger values, on average, than the raw values 
$x_{ij}$. Therefore, the $SSE_{Var}$ criterion naturally receives greater weight 
when computing the k-plus criterion than the standard $SSE$ criterion. This 
imbalance is unwarranted because optimizing similarity with regard to variances 
is not more important than optimizing similarity with regard to means. A simple 
solution for this issue is to standardize all values $x_{ij}$ and $y_{ij}$ to 
the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

Optimizing a bicriterion objective such as $SSE_{kplus}$ may not succeed as one 
may hope. The ideal outcome would be to find a partition where both the means 
and the variances are very similar between groups. However, research on 
multi-objective optimization generally shows that cases are rare where a single 
solution satisfies all criteria to a satisfactory degree. Usually, a tradeoff is 
inevitable. It is an open question whether k-plus anticlustering is a expection 
from this rule. Next, a motivating example illustrates that in anticlustering, 
aiming to simultaneously optimize similarity of means and variances is not only 
desirable, but may actually be feasible. This notion is later confirmed when 
applying k-plus anticlustering in simulation studies and examples using real 
data.

For this initial example, I created 14 data points from a univariate normal 
distribution ($M = 15$, $SD = 3$) and generated all `r n_partitions(14, 2)` 
possible ways to partition the 14 data points into $K = 2$ equal-sized groups. 
For each partition, the criteria SSE and $SSE_{Var}$ were computed. Figure 1 
shows the scatterplot of these criteria across partitions. Note that for both 
criteria, a larger value is better, i.e., indicates higher similarity with 
regard to means and variances, respectively. Ideally, to obtain overall 
between-group similarity, a partition is chosen that minimizes the difference 
with regard to both means and variances (see the triangular symbol). However, 
optimal k-means anticlustering would create two groups that are not particularly 
similar with regard to their variance. The partition that is represented by the 
squared symbol has the overall best highest SSE, but is outperformed by many 
other partitions with regard to $SSE_{Var}$. The diamond symbol represents 
that partition that where the two groups have the least difference in 
variances.

Next: use the partition that optimizes $SSE_{kplus}$!

@@

```{r, fig.cap = "The relationship between the two criteria SSE and $SSE_{Var}$ across partitions is shown. The partition represented by the diamond minimizes the difference in variances between groups; the partition represented by the square minimizes the difference in means between groups; the partition represented by the triangular is a good tradeoff, satisfying both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "SSE",
  ylab = "SSE_Var"
)
